MSD panels 1, 2 and 3 Explant Supernatant

### 2016-09-22 MSD_panels_1_2_3_explant_sup_analysis.Rmd
edit final plot
Making a heat map of logFC 3hr vs 24hrs

explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds %>%
mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.01, Estimated_Stock_Conc))

@@

@@

```{r  subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}

@@

mock <- avgConc_explant_all_panels_noStds_AdjBelow %>%
  filter(Virus == "M")

@@

no_mock_avgConc_explant_all_panels_noStds_AdjBelow <-
  avgConc_explant_all_panels_noStds_AdjBelow %>%
  filter(Virus !="M")

@@

#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale. So I'm setting all zeros in the minusMockEstConc column to 0.01.

#Also, in cases where the Mock values are > Virus values, we would get a negative number for the result and not be able to plot it on a log scale. So I will take the log of each, then subtract, then plot on a linear scale, so I can have the negatives there.

#Recall that log(a/b) = log(a)-log(b)
#so if Mock = 100 and V = 10, log 100/10 = log (0.1) = -1
# and log(100)-log(10) = 1-2 = -1

mergeMock <- mergeMock %>%
  mutate(logMock = log10(avgEstConcMock))%>%
  mutate(logVirus = log10(avgEstConcVirus))%>%
  mutate(MockSubtracted = logVirus - logMock)

```{r make adjustments for nice plotting,echo=FALSE}

mergeMock<-mergeMock %>%
  mutate(Timepoint = ifelse(Timepoint == "T1", "3",
                            ifelse(Timepoint=="T2","8","24")))

mergeMock<-mergeMock%>%
  mutate(Virus = ifelse(Virus == "V1", "SD90",
                        ifelse(Virus == "V2", "186","Mock")))

mergeMock$Timepoint<-factor(mergeMock$Timepoint,levels = c("3","8","24"))

@@

```{r  individual analyte plots, fig.width = 6,fig.height= 13, echo=FALSE}

ggplot(mergeMock, aes(x = Timepoint, y = MockSubtracted))+
   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
   labs( y = "log 10 Estimated Stock Concentration pg/ml")+
   facet_grid(Assay~Virus, scales = "free_x")+
   theme(axis.text.x = element_text(size = 11))+
  #geom_hline(yintercept = 0)+
  ggtitle("Mock Subtracted Analyte Concentrations")

@@

```{r  subset plots, fig.width = 6,fig.height= 13, echo=FALSE}

#Make plots just showing data for donors 317, 318, 323,324

subset<-mergeMock%>%
  filter(Donor == "317"|Donor == "318"|Donor =="323"|Donor =="324")

#plot it

ggplot(subset, aes(x = Timepoint, y = MockSubtracted))+
   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
   labs( y = "log10 Estimated Stock Concentration pg/ml")+
   facet_grid(Assay~Virus, scales = "free_x")+
   theme(axis.text.x = element_text(size = 11))+
  ggtitle("Mock Subtracted Analyte Concentrations")

>>>>

explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds %>%
mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.001, Estimated_Stock_Conc))

@@

```{r make adjustments for nice plotting, echo=FALSE, message = FALSE, warning = FALSE}

explant_all_panels_noStds_AdjBelow<-explant_all_panels_noStds_AdjBelow %>%
  mutate(Timepoint = ifelse(Timepoint == "T1", "3",
                            ifelse(Timepoint=="T2","8","24")))

#factor to get order right
explant_all_panels_noStds_AdjBelow$Timepoint<-factor(explant_all_panels_noStds_AdjBelow$Timepoint,levels = c("3","8","24"))

explant_all_panels_noStds_AdjBelow<-explant_all_panels_noStds_AdjBelow%>%
  mutate(Virus = ifelse(Virus == "V1", "SD90",
                        ifelse(Virus == "V2", "186","Mock")))

explant_all_panels_noStds_AdjBelow$Virus<-factor(explant_all_panels_noStds_AdjBelow$Virus,levels = c("Mock","SD90","186"))

@@

```{r  merge and subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}

@@

mock <- avgConc_explant_all_panels_noStds_AdjBelow %>%
  filter(Virus == "Mock")

@@

no_mock_avgConc_explant_all_panels_noStds_AdjBelow <-
  avgConc_explant_all_panels_noStds_AdjBelow %>%
  filter(Virus !="Mock")

@@

mergeMock <- mergeMock %>%
  mutate(mockSubtracted = avgEstConcVirus - avgEstConcMock)

@@

```{r setup for individual analyte HEATMAP, echo=FALSE}

#I want to make a heatmap that has the following data

#NO 8hr timepoint data

#tiles represent foldchange of Mock subtracted 24hr data over mock-subtracted 3hr. So if 24hr>3hr, you get a ratio <1 and if 24hr <3hr you get a ratio >1

#Want gradient intensity to max out at FC of 10.

#Exclude 8hr data.

no8mergeMock <- mergeMock %>%
  filter(Timepoint != "8")

# Extract the 3hr data and select a few columns

timepoint3<- no8mergeMock %>%
  filter(Timepoint == "3")%>%
  select(Donor, Assay, Sample, Virus, mockSubtracted)

#extract 24 hr data and select a few columns
timepoint24<- no8mergeMock %>%
  filter(Timepoint =="24")%>%
  select(Donor, Assay, Sample, Virus, mockSubtracted)

```{r merge 24 and 3, check NAs, echo=FALSE}
#merge 24hr and 3hr data

merge24And3<- merge(timepoint24,timepoint3, by  = c("Donor","Assay","Virus"), all.x = TRUE, all.y = TRUE)

#There are 100 rows where we have data for 24hr but NOT for 3hr.
which24no3<-merge24And3[which(is.na(merge24And3$Sample.y)),]

#there are NO rows where we have data for 3hr but not for 24hr
which3no24<-merge24And3[which(is.na(merge24And3$Sample.x)),]

#remove the NA's for tidyness

merge24And3 <- na.omit(merge24And3)

#Fix up colnames      
names(merge24And3)[4:7]<- c("Sample.24", "mockSubtracted.24","Sample.3", "mockSubtracted.3")
```

```{r adjust zeros, check NAs, echo=FALSE}

#Problems with zeros:
#If data from both timepoints is zero we'll get NaN for the ratio

#If data from time 3 = 0, we'll get Inf or -Inf
#If data from time 24 = zero, we'll get zero

#I'm going to change all zeros to 0.001 to avoid these problems #I think this is ok since a heatmap isnt really quantitative

adjZerosMerge24And3<- merge24And3 %>%
  mutate(mockSubtracted.24 = ifelse(mockSubtracted.24 == 0, 0.001,mockSubtracted.24))%>%
  mutate(mockSubtracted.3 = ifelse(mockSubtracted.3 == 0, 0.001,mockSubtracted.3))

```{r get ratio, check NAs, echo=FALSE}

#Do the division to get a foldchange. Since we are looking at ratios, it doesnt give any info if the number is negative, plus we can't show that on a log scale. 
#So I'm taking the abs value of the ratio.
#and also making a column for the log10 of the ratio for plotting.

adjZerosMerge24And3 <- adjZerosMerge24And3 %>%
  mutate(AbsVal_mockSub24_Over_mockSub3 = abs(mockSubtracted.24/mockSubtracted.3))%>%
  mutate(log10ratio = log10(AbsVal_mockSub24_Over_mockSub3))

```{r logFC heatmap, check NAs, echo=FALSE}

ggplot(adjZerosMerge24And3, aes(x = Donor, y = Assay))+
  geom_tile(aes(fill = log10ratio), color = "white")+
  scale_fill_gradient(low = "green", high = "red")+
  facet_wrap(~Virus)+
  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")

@@
